\documentclass[12pt]{artikel3}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}

\newcommand{\R}{\mathbb{R}}
\newcommand{\bigboxtimes}{\boxtimes}

\title{Alternating Least Squares using $QR$-factorizations}
\author{Jan Van lent}

\begin{document}

\maketitle

Suppose we are looking for a separated vector of the form
\[x = \sum_{l=1}^r s_l \, x^1_l \otimes \dotsb \otimes x^d_l\]
or for each component
\[x_{j_1,\dotsc,j_d} = 
\sum_{l=1}^r s_l \, x^1_{j_1,l} \dotsm x^d_{j_d,l}.\] 
To emphasize that this representation consists of a vector \(s =
[s_l] \in \R^r\) and matrices \(X^i = [x^i_{j_i,l}] \in \R^{M_i, r}\), 
I introduce the notation
\[x = ( X^1 \boxtimes \dotsb \boxtimes X^d ) s,\]
where \(A \boxtimes B\) is the matrix with as columns the tensor
product of the corresponding columns of \(A\) and \(B\), i.e.,
\[ A \boxtimes B = 
\left[\begin{array}{ccc} a_1 & \cdots & a_r \end{array}\right] \boxtimes 
\left[\begin{array}{ccc} b_1 & \cdots & b_r \end{array}\right] =
\left[\begin{array}{ccc} a_1 \otimes b_1 & \cdots & 
a_r \otimes b_r \end{array}\right].\]

Let the vector to be approximated be 
\[y = ( Y^1 \boxtimes \dotsb \boxtimes Y^d ) t,\]
where \(t \in \R^q\) and \(Y^i \in \R^{M_i, q}\).  

The alternating least squares algorithm updates \(s\) and \(X^1\),
then \(s\) and \(X^2\), etc.

Suppose \(d = 2\) and we want to find \(X^2\) and \(s\) with \(X^1\)
fixed. The system of equations is
\[ ( X^1 \boxtimes X^2 ) s \approx ( Y^1 \boxtimes Y^2 ) t. \]
Let the reduced \(QR\)-factorization of \(X^1\) be
\[X^1 = Q^1 R^1,\]
where \(R^1 \in \R^{r \times r}\) is lower triangular and
\(Q^1 \in \R^{M_1 \times r}\) has orthogonal columns (see
\cite{GVL96}).  Using this factorization the system of \(M_1\)
equations can be transformed into
\[ ( R^1 \boxtimes X^2 ) s = ( \tilde{Y}^1 \boxtimes Y^2 ) t, \]
where \(\tilde{Y}^1 = Q^T Y^1 \in \R^{r \times r}\). The
solution of this system of \(r\) equations is the least squares
solution we are looking for. The last column of \(X^2\) and the last
element of \(s\) are found by normalizing the vector
\[ ( \tilde{Y}^1_{r,*} \boxtimes Y^1 ) t / R_{r,r}. \]
Using these values the second to last column can be found, etc. This
is essentially back substitution with \(R^1\) and multiple right hand
sides. This can also be seen by introducing the matrices \(Z = (X^2
s)^T\) so that left hand side of the original system becomes \(X^1
Z\). The matrix \(Z\) is the solution of a standard least squares
problem with multiple right hand sides.

The extension to \(d > 2\) is a bit involved, but possible. Given
factorizations \(X^i = Q^i R^i\) for \(i \neq k\), the system of
equations
\[ ( X^1 \boxtimes \dotsb \boxtimes X^k \boxtimes \dotsb \boxtimes X^d ) s 
\approx
( Y^1 \boxtimes \dotsb \boxtimes Y^k 
\boxtimes \dotsb \boxtimes Y^d ) t,\]
becomes
\[ ( R^1 \boxtimes \dotsb \boxtimes X^k \boxtimes \dotsb \boxtimes R^d ) s 
\approx
( \tilde{Y}^1 \boxtimes \dotsb \boxtimes Y^k 
\boxtimes \dotsb \boxtimes \tilde{Y}^d ) t.\]
The factorizations for \(M_i = M\) take about \(d (2Mr^2 - \frac{2}{3}
r^3)\) operations. Including the transformations of \(Y^i\) to
\(\tilde{Y}^i\) adds about another \(d(Mq^2+\frac{2}{3}q^3)\)
operations.  The last column of \(X^k\) can be found by considering
all the rows of \(\bigboxtimes_{i \neq k} R^i\) that have only one
non-zero element (the last). Using this information the second to last
column is found by considering all the rows having two non-zero
elements, etc. Each column of \(X^k\) is the solution of a least
squares problem with a column vector as matrix.  The same solution can
also be found using a further \(QR\)-factorization of
\(\bigboxtimes_{i \neq k} R^i\). The reduction to lower triangular
form can exploit the structure of this matrix. We can, for example,
first zero out \(R^d\) by working on \(R^{d-1} \boxtimes R^d\), then
\(R^{d-1}\), and so on upto \(R^2\). I think transforming
\(\bigboxtimes_{i \neq k} R^i\) and \(\bigboxtimes_{i \neq k}
\tilde{Y}^i\) takes about \(d(\frac{2}{3} r^3 + 2 q r^2)\) operations.

For a standard least squares problem
\[Ax \approx b,\]
regularization of the normal equations takes the form
\[(A^T A + \nu I) x = A^T b.\]
Regularization for the QR, SVD or MGS approach is obtained by solving
\[\left[\begin{array}{c}A \\ \sqrt{\nu} I\end{array}\right]
\approx
\left[\begin{array}{c}b \\ 0\end{array}\right],\]
see for example p.~47 of \cite{Kel99}. I do not immediately see how to
do the equivalent for the structured systems above, but maybe
appending a scaled identity matrix to each \(X^i\) would be
appropriate (maybe scaled by \(\nu^{\frac{1}{2d}}?\) or \(\mu\)).

An approach based on \(QR\) is more expensive than the approach based
on normal equations, but, provided regularization can be introduced, it
may be able to obtain more accurate results.

I have not put any thought into finding a \(QR\) approach for the
least squares problem \(A x \approx y\), involving a matrix \(A\) and
vectors \(x\) and \(y\) in separated representation, but I think that
this should also be possible.

\bibliographystyle{alpha}
\bibliography{qr}

\end{document}
